---
title: "Suspected drug deaths in Scotland: January to March 2023"

output: 
  word_document:
    reference_docx: DRDs_briefing_template.docx
  
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      dpi = 300)

```

```{r README}

##Producing this briefing requires reading in several files:
#1) Functions - main formatting and tidying of data
source("Mainfunctions/1_Functions.R")

#2) Wrangling - prepares the data for analysis
source("Mainfunctions/2_Wrangling.R")

#3) Misc_functions - specifically pulls out data for the most recent quarter, and the most recent 12 months
source("Mainfunctions/3_MISC_functions.R")

#############################
#############################
#############################
#############################
#############################

#FINALLY - the relevant R script for whichever quarter you're producing the briefing in - THIS IS THE ONLY ONE YOU CHANGE

source("Quarterlyfunctions/Q1_Functions.R")


```

## 1. Background

The Scottish Government has a National Mission to reduce drug deaths and improve the lives of those affected by drugs. This quarterly report presents Police Scotland management information to provide an indication of current trends in suspected drug deaths in Scotland.

## 2. Summary

There were `r latest_1Q` suspected drug deaths between January and `r last_month` `r year`. This was `r latest_fig_prevQ_prop`% (`r latest_fig_diff`) `r latest_which2` than the previous quarter and `r latest_1Q_comparison_prop`% (`r latest_1Q_comparison_diff`) `r latest_which2` than during the same period of `r year_prev`. Following a downward trend in the rolling 12-month suspected drug death total through Q3 2022, the rolling number of suspected drug deaths has risen slightly over the last two quarters (Figure 1).

## 3. Main points

```{r figure1, include=T, fig.dim = c(7,5)}
#Derived from the function in 1_Functions.R

PoliceDivision %>% 
  Plot_fun()
```

Between January and `r last_month` `r year`:

  * There were `r latest_1Q` suspected drug deaths, `r latest_1Q_comparison_prop`% (`r latest_1Q_comparison_diff`) `r latest_which2` than during the same period of `r year_prev`.
  * Males accounted for `r Males_latest`% of suspected drug deaths. This compares with `r Males_prev`% between January and `r last_month` `r year_prev`;
  * There were `r Females` suspected drug deaths of females, `r Females_which` of `r Females_prop`% (`r Females_prev`) compared to the same period in `r year_prev`;
  * A majority (`r Age_prop_middleaged`%) of suspected drug deaths were of people aged between 35 and 54. This is broadly in line with previous quarters;
  * There were `r Age_young` suspected drug deaths in the under 25 age group, `r Age_young_prop`% (`r Age_young_comparison`) `r Age_which` than between January and `r last_month` `r year_prev`;
  * The Police Divisions with the greatest number of suspected drug deaths were: `r Ranking_PoliceDiv`.
  

  **Note that numbers of suspected drug deaths fluctuate from quarter to quarter. Care should be taken not to interpret movements between individual calendar quarters as indicative of any long term trend.**

  * There were `r rolling12months` suspected drug deaths over the 12 months to `r last_month` `r year`, `r rolling12months_propchange`% (`r rolling12months_absolutechange`) `r rolling12months_which` than the 12 months to `r last_month` `r year_prev`. 

More detail on the statistics presented above is available in the workbook presented alongside this report [LINK].

## 4. Methodology

The data in this report is operational information from Police Scotland who compile figures on the basis of reports from police officers attending scenes of death. This is management information and not subject to the same level of validation and quality assurance as Official Statistics. National Statistics on drug deaths in Scotland are published annually by [National Records of Scotland](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/drug-related-deaths-in-scotland/2021). 

More information on definitions, data sources, methodology and interpretation of the data, including the historical relationship between suspected drug death figures and drug misuse death statistics, can be found in the Methodology Annex[LINK].

## 5. Next update

The next RADAR report, containing more up-to-date suspected drug death figures, will be publisehd in July 2023.

Public Health Scotland (PHS) compile a [Rapid Action Drug Alerts and Response (RADAR) Quarterly report](https://publichealthscotland.scot/publications/rapid-action-drug-alerts-and-response-radar-quarterly-report/rapid-action-drug-alerts-and-response-radar-quarterly-report-january-2023/#section-2) of drug-related indicators in order to inform action to prevent drug harms and deaths. In January 2023, PHS added suspected drug deaths to their set of RADAR indicators, presenting weekly data on suspected drug deaths from Police Scotland. PHS will now publish data on this indicator quarterly, with the next publication due for release in April 2023. In light of this, Scottish Government, together with partners, will consider the future of the ‘Suspected drug deaths in Scotland’ report.

The next annual drug misuse deaths National Statistics publication from NRS is due to be released in Summer 2023.

## 6. Accessing help and support for problem drug use
If you or anyone you know is affected by drug use, support is available via the following organisations:

  * [Scottish Drugs Forum](https://www.sdf.org.uk/)
  * [Scottish Families Affected by Alcohol & Drugs](https://www.sfad.org.uk/service-directory)
  * [Know the Score](https://www.knowthescore.info/)
  * [We Are With You](https://www.wearewithyou.org.uk/)

  
### Correspondence and enquiries

For enquiries about this publication please contact:

Population Health Analysis Unit  
Health & Social Care Analysis  
E-mail: [HSCAnalysisHub@gov.scot](HSCAnalysisHub@gov.scot)  

For general enquiries about Scottish Government statistics please contact:   
Office of the Chief Statistician  
E-mail: [statistics.enquiries@gov.scot](statistics.enquiries@gov.scot)  

Scottish Government welcomes feedback and suggestions on statistical products. Please complete the [feedback survey](https://consult.gov.scot/digital-communications/95832709/) to share your views.